---
title: "Methods & code"
subtitle: "How field measurements become numbers: the two models behind most of my work, running live on this page."
title-block-banner: images/field/banner-methods.jpg
title-block-banner-color: white
toc: true
toc-location: left
---

```{r}
#| label: setup
#| include: false
library(dplyr); library(tidyr); library(readr); library(plotly)

pal <- c(deep = "#0e3b43", mid = "#1d7a73", moss = "#6e8c3a", ochre = "#b08b3e")
```

Two measurements carry most of what I do in the field: a sonde logging oxygen,
pH and conductivity every few minutes, and a pulse of solute released at the
top of a reach. Neither is interesting on its own. What makes them useful is
the mass balance you write around them, and this page walks through those two
balances. Every figure below is produced by the code beside it when the site
is built.

## 1. Reading metabolism off a diel oxygen curve

A stream gains oxygen by photosynthesis and by exchange with the atmosphere,
and loses it to respiration and, when supersaturated, back to the atmosphere.
Over a reach deep $z$, that is one equation:

$$
\frac{dO_2}{dt} \;=\; \underbrace{\frac{GPP \cdot f_{PAR}(t)}{z}}_{\text{photosynthesis}}
\;-\; \underbrace{\frac{ER}{z}}_{\text{respiration}}
\;+\; \underbrace{k\,\bigl(O_{2,sat}(T) - O_2\bigr)}_{\text{reaeration}}
$$

Gross primary production follows light, so it appears only during the day and
is distributed through it by the normalised light curve $f_{PAR}$. Ecosystem
respiration runs day and night. Reaeration is the term that hides the other
two: a well-mixed riffle exchanges gas so fast that biology barely moves the
oxygen signal, while a slow pool amplifies the same metabolism into a wide
daily swing. Estimating $k$ badly is the fastest way to get metabolism wrong.

The figure below runs that equation forward. Drag the sliders: the daily
amplitude responds to GPP and ER, and collapses as $k$ grows.

```{r}
#| label: metab-grid
#| code-summary: "Forward metabolism model"
o2_sat <- function(temp = 25) 14.652 - 0.41022 * temp +
  0.007991 * temp^2 - 0.000077774 * temp^3

diel_o2 <- function(gpp, er, k600, z = 0.25, temp = 25, dt = 1 / 96) {
  t <- seq(0, 1, by = dt)                       # fração do dia
  light <- pmax(0, sin(pi * (t - 0.25) / 0.5))  # nascer 6h, pôr 18h
  light[t < 0.25 | t > 0.75] <- 0
  f_par <- light / (sum(light) * dt)            # integra 1 no dia
  sat <- o2_sat(temp)
  o2 <- numeric(length(t)); o2[1] <- sat
  for (i in 2:length(t)) {
    d <- (gpp * f_par[i] - er) / z + k600 * (sat - o2[i - 1])
    o2[i] <- o2[i - 1] + d * dt
  }
  tibble(hour = t * 24, o2 = o2, gpp = gpp, er = er, k600 = k600)
}

grid <- expand_grid(gpp = seq(2, 14, by = 2), er = seq(2, 14, by = 2),
                    k600 = c(2, 6, 12, 24))
# dt maior aqui só para o grid: 49 pontos por combinação mantêm a forma da
# curva e o JSON enviado ao navegador em poucas centenas de kB.
metab <- purrr::pmap_dfr(grid, function(gpp, er, k600)
  diel_o2(gpp, er, k600, dt = 1 / 48))

ojs_define(metab_raw = metab)
```

```{ojs}
//| echo: false
metab = transpose(metab_raw)

viewof gpp_in = Inputs.range([2, 14], {step: 2, value: 8,
  label: "GPP (g O₂ m⁻² d⁻¹)"})
viewof er_in = Inputs.range([2, 14], {step: 2, value: 8,
  label: "ER (g O₂ m⁻² d⁻¹)"})
viewof k_in = Inputs.select([2, 6, 12, 24], {value: 6,
  label: "k₆₀₀ (d⁻¹)"})

sel = metab.filter(d => d.gpp === gpp_in && d.er === er_in && d.k600 === k_in)

Plot.plot({
  height: 300,
  marginLeft: 55,
  x: {label: "Hour of day", domain: [0, 24], ticks: [0, 6, 12, 18, 24]},
  y: {label: "O₂ (mg L⁻¹)", grid: true},
  marks: [
    Plot.ruleY([8.24], {stroke: "#b08b3e", strokeDasharray: "4 3"}),
    Plot.lineY(sel, {x: "hour", y: "o2", stroke: "#1d7a73", strokeWidth: 2.5})
  ]
})
```

The dashed line is saturation at 25 °C. When GPP and ER are equal the curve
returns to where it started after 24 h and the reach is metabolically
balanced; when respiration wins, the whole curve sits below saturation and the
stream is a net source of CO~2~ to the atmosphere.

## 2. Why the carbonate system moves with the oxygen

In a travertine stream the same photosynthesis that produces oxygen consumes
dissolved CO~2~. Removing CO~2~ at fixed alkalinity raises pH, which converts
bicarbonate into carbonate, which raises the calcite saturation state

$$\Omega = \frac{[\mathrm{Ca^{2+}}][\mathrm{CO_3^{2-}}]}{K_{sp}}$$

and once $\Omega$ is comfortably above 1, calcite precipitates and calcium
leaves the water. Respiration at night does the reverse. The consequence is
that oxygen, pH and calcium in these streams are three views of one daily
cycle, and an oxygen record alone cannot tell you how much of the carbonate
signal is biological.

The model below is the same mass balance as above with carbonate chemistry
attached: dissolved inorganic carbon responds to metabolism and to CO~2~
exchange with the air, pH is solved from DIC and alkalinity at each step, and
calcite precipitates at a rate proportional to $(\Omega - 1)$, drawing down
calcium and alkalinity.

```{r}
#| label: carbonate-model
#| code-summary: "Coupled metabolism and carbonate model"
K1 <- 10^-6.35; K2 <- 10^-10.33; Kw <- 1e-14; Ksp <- 10^-8.48

# pH a partir de DIC e alcalinidade, por bisseção em [H+]
solve_h <- function(dic, alk) {
  f <- function(h) {
    denom <- h^2 + K1 * h + K1 * K2
    hco3 <- dic * K1 * h / denom
    co3  <- dic * K1 * K2 / denom
    hco3 + 2 * co3 + Kw / h - h - alk
  }
  lo <- 1e-12; hi <- 1e-4
  for (i in 1:60) {
    mid <- sqrt(lo * hi)
    if (f(mid) > 0) lo <- mid else hi <- mid
  }
  sqrt(lo * hi)
}

diel_carb <- function(gpp, er = 8, z = 0.25, k600 = 6,
                      alk0 = 4.5e-3, ca0 = 2.2e-3, dic0 = 4.8e-3,
                      k_cal = 6, dt = 1 / 288) {
  t <- seq(0, 1, by = dt)
  light <- pmax(0, sin(pi * (t - 0.25) / 0.5)); light[t < 0.25 | t > 0.75] <- 0
  f_par <- light / (sum(light) * dt)
  sat <- o2_sat(25); co2_air <- 1.2e-5          # mol/L em equilíbrio com o ar

  o2 <- dic <- alk <- ca <- ph <- omega <- numeric(length(t))
  o2[1] <- sat; dic[1] <- dic0; alk[1] <- alk0; ca[1] <- ca0

  for (i in seq_along(t)) {
    if (i > 1) {
      # metabolismo em mol C por litro por dia (32 g O2 = 1 mol)
      p <- gpp * f_par[i] / (z * 32 * 1000)
      r <- er / (z * 32 * 1000)
      h <- solve_h(dic[i - 1], alk[i - 1])
      co2 <- dic[i - 1] * h^2 / (h^2 + K1 * h + K1 * K2)
      co3 <- dic[i - 1] * K1 * K2 / (h^2 + K1 * h + K1 * K2)
      om  <- ca[i - 1] * co3 / Ksp
      prec <- k_cal * max(0, om - 1) * 1e-4      # mol CaCO3 L-1 d-1

      dic[i] <- dic[i - 1] + (r - p - prec - k600 * (co2 - co2_air)) * dt
      alk[i] <- alk[i - 1] - 2 * prec * dt
      ca[i]  <- ca[i - 1] - prec * dt
      o2[i]  <- o2[i - 1] + ((gpp * f_par[i] - er) / z + k600 * (sat - o2[i - 1])) * dt
    }
    h <- solve_h(dic[i], alk[i])
    ph[i] <- -log10(h)
    co3 <- dic[i] * K1 * K2 / (h^2 + K1 * h + K1 * K2)
    omega[i] <- ca[i] * co3 / Ksp
  }
  tibble(hour = t * 24, gpp = gpp, o2 = o2, ph = ph,
         ca = ca * 1000, omega = omega)          # Ca em mmol L-1
}

carb <- purrr::map_dfr(seq(2, 14, by = 2), diel_carb)
ojs_define(carb_raw = carb)
```

```{ojs}
//| echo: false
carb = transpose(carb_raw)

viewof gpp_carb = Inputs.range([2, 14], {step: 2, value: 10,
  label: "GPP (g O₂ m⁻² d⁻¹)"})

csel = carb.filter(d => d.gpp === gpp_carb)

panel = (y, label, colour) => Plot.plot({
  height: 190,
  marginLeft: 60,
  x: {label: "Hour of day", domain: [0, 24], ticks: [0, 6, 12, 18, 24]},
  y: {label: label, grid: true},
  marks: [Plot.lineY(csel, {x: "hour", y: y, stroke: colour, strokeWidth: 2.5})]
})

html`<div>
  ${panel("o2", "O₂ (mg L⁻¹)", "#1d7a73")}
  ${panel("ph", "pH", "#0e3b43")}
  ${panel("ca", "Ca²⁺ (mmol L⁻¹)", "#b08b3e")}
</div>`
```

Raise GPP and the three panels move together: oxygen and pH peak in the
afternoon, calcium falls through the day as calcite forms, and the overnight
recovery is only partial, because the calcium that precipitated does not come
back. That asymmetry is what a travertine reach records as rock.

::: {.callout-note collapse="true"}
## What this model is and is not

Illustrative, with realistic but generic parameters: 25 °C, fixed
temperature, one reach depth, a single first-order precipitation rate. It
reproduces the coupling, not any particular stream. Field estimates require
measured light, temperature, reaeration and alkalinity, and are fitted rather
than assumed.
:::

## 3. Nutrient uptake from a slug addition

Release a pulse of salt and nutrient at the top of a reach and sample at the
bottom. The salt is conservative: whatever passes the downstream station is
what was added, spread in time by dispersion and transient storage. The
nutrient rides the same hydrology, so any nutrient missing relative to the
salt was taken up by the streambed between the two stations.

```{r}
#| label: fig-btc
#| fig-cap: "Paired breakthrough curves from one injection. Synthetic example data."
#| code-summary: "Breakthrough curves"
btc <- read_csv("data/example_btc.csv", show_col_types = FALSE) |>
  mutate(cond_bg = cond_uS_cm - median(cond_uS_cm[minutes < 5]),
         nh4_bg  = nh4_ugL  - median(nh4_ugL[minutes < 5]))

plot_ly(btc, x = ~minutes) |>
  add_lines(y = ~cond_bg, name = "Conservative tracer",
            line = list(color = pal[["deep"]], width = 2.5)) |>
  add_lines(y = ~nh4_bg, name = "NH₄⁺", yaxis = "y2",
            line = list(color = pal[["moss"]], width = 2.5)) |>
  layout(xaxis = list(title = "Minutes since injection"),
         yaxis = list(title = "Conductivity above background (µS cm⁻¹)"),
         yaxis2 = list(title = "NH₄⁺ above background (µg L⁻¹)",
                       overlaying = "y", side = "right", showgrid = FALSE),
         legend = list(orientation = "h", y = 1.1),
         hovermode = "x unified")
```

Comparing the two curves point by point gives the uptake rate. Scaled by
velocity and depth it becomes the quantity that lets reaches of different size
be compared, the uptake length

$$S_w = \frac{u \, h}{V_f}$$

the average distance a nutrient molecule travels before being taken up. Fast
water with a slow biofilm carries nutrients far downstream; a shallow, active
reach retains them within metres.

```{ojs}
//| echo: false
viewof u = Inputs.range([0.02, 0.6], {step: 0.01, value: 0.15,
  label: "Velocity u (m s⁻¹)"})
viewof h = Inputs.range([0.05, 0.8], {step: 0.05, value: 0.25,
  label: "Depth h (m)"})
viewof vf = Inputs.range([0.5, 20], {step: 0.5, value: 5,
  label: "Uptake velocity V_f (mm min⁻¹)"})

sw = u * h / (vf / 1000 / 60)

html`<p style="font-size:1.1rem;margin-top:0.5rem">
Uptake length <b>S<sub>w</sub></b> =
<b style="color:#1d7a73">${sw.toFixed(0)} m</b>
 · a molecule travels that far before it is taken up.
</p>`
```

## Reproducibility

Every number and figure on this page is computed when the site is built, from
the code folded into each block. Nothing is a screenshot. The same discipline
applies to the analyses behind the papers: raw data in, figures out, one
command.

```{r}
#| label: session
#| code-summary: "Session info"
sessioninfo::session_info(pkgs = "attached")
```
