---
title: "Publications"
subtitle: "Peer-reviewed articles and research software."
title-block-banner: images/field/banner-publications.jpg
title-block-banner-color: white
body-classes: banner-md banner-center
toc: false
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(readr)
library(glue)
library(htmltools)
```

Talks, posters and work in preparation are on the [CV](cv.qmd).

```{r}
#| label: pub-list
#| code-summary: "Render the list from CSV"
#| output: asis
pubs <- read_csv("data/publications.csv", show_col_types = FALSE) |>
  arrange(desc(year), type, title)

render_group <- function(df, heading) {
  if (nrow(df) == 0) return(invisible(NULL))
  cat("\n## ", heading, "\n\n", sep = "")
  for (i in seq_len(nrow(df))) {
    r <- df[i, ]
    link <- if (!is.na(r$link) && nzchar(r$link)) {
      glue("[{r$link_label}]({r$link}){{.tag}}")
    } else ""
    cat(glue(
      "::: {{.line}}\n",
      "### {r$title}\n",
      "{r$authors}. *{r$venue}* ({r$year}).\n\n",
      "[{r$status}]{{.tag}} {link}\n",
      ":::\n\n"
    ))
  }
}

render_group(filter(pubs, type == "Article"),  "Peer-reviewed articles")
```

## Software

### StreamLog

A single-file HTML logger for multiparameter sonde readings in the field:
range checks as you type, CSV export, and no dependency on signal, battery-
hungry apps or a laptop. It runs offline in any browser, including a phone
held over a stream.

[Open in a new tab](https://kauanfonseca.github.io/streamlog/){.tag}
[Source code](https://github.com/kauanfonseca/streamlog){.tag}
[10.5281/zenodo.22784736](https://doi.org/10.5281/zenodo.22784736){.tag}
[MIT licence]{.tag}

::: {.app-embed}
<iframe src="https://kauanfonseca.github.io/streamlog/" loading="lazy"
        title="StreamLog running live"></iframe>
:::

The version above is live, not a screenshot: type a reading and it validates.
Nothing you enter here is stored anywhere.
